---
title: "Worked example: bounds for a differentially misclassified mediator (gestational hypertension)"
author: "Davood Tofighi, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: cosmo
    highlight-style: github
    df-print: paged
execute:
  eval: true
  echo: true
vignette: >
  %\VignetteIndexEntry{Worked example: bounds for a differentially misclassified mediator (gestational hypertension)}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## The question

Does **gestational hypertension** mediate the association between **advanced maternal
age** and **preterm birth**? Gestational hypertension recorded on the birth certificate is
a *differentially misclassified* binary mediator: its sensitivity is poor and reporting
accuracy plausibly depends on the outcome. With no validation data, the natural direct
(NDE) and indirect (NIE) effects are not point-identified --- but `medrobust` returns a
**sharp identified set** over a transparent sensitivity region, plus an Imbens--Manski
confidence interval.

This is the same substantive application used by the closest prior *point*-identification
method (Hochstedler Webb and Wells, 2025); here we instead bound the effects without an
auxiliary identifying assumption.

```{r setup}
library(medrobust)
data("gesthtn")
# A: advanced maternal age (>=35) | M_star: birth-cert gestational HTN (misclassified)
# Y: preterm birth (<37 wk)       | C1: parity (any prior live birth)
table(M_star = gesthtn$M_star, Y = gesthtn$Y)
```

## A naive analysis (treats the surrogate as truth)

```{r naive}
fM <- glm(M_star ~ A + C1, data = gesthtn, family = binomial())
fY <- glm(Y ~ A + M_star + C1, data = gesthtn, family = binomial())
naive_NDE_OR <- exp(coef(fY)[["A"]])
round(naive_NDE_OR, 3)
```

## The sensitivity region

We ground the region in a birth-certificate validation study (Dietz et al., 2015): the
gestational-hypertension item has *poor* sensitivity (`< 70%`) and *excellent* specificity
(`> 90%`); differential sensitivity ($\psi_{Sn} > 1$) is plausible because preterm,
higher-acuity deliveries receive closer chart abstraction.

```{r region}
region <- sensitivity_region(
  sn0_range    = c(0.50, 0.70),   # poor baseline sensitivity
  sp0_range    = c(0.90, 0.99),   # excellent baseline specificity
  psi_sn_range = c(1.0, 3.0),     # non-differential -> moderate differential
  psi_sp_range = c(1.0, 1.0)      # specificity non-differential
)
```

## Partial-identification bounds + Imbens--Manski CI

```{r bounds}
b <- bound_ne(
  data = gesthtn, exposure = "A", mediator = "M_star", outcome = "Y",
  confounders = "C1", misclassified_variable = "mediator",
  sensitivity_region = region, effect_scale = "OR",
  ci_method = "analytic", ci_n_boot = 100, verbose = FALSE
)

# Identified set [L, U] for each effect (OR scale)
c(NDE_lower = b@NDE_lower, NDE_upper = b@NDE_upper)
c(NIE_lower = b@NIE_lower, NIE_upper = b@NIE_upper)

# Imbens-Manski CI for the partial-identification set
b@analytic_ci$NDE[c("ci_lower", "ci_upper")]
b@analytic_ci$NIE[c("ci_lower", "ci_upper")]
```

## Reading the result

The indirect-effect identified set summarizes how the mediated effect could range as the
misclassification mechanism varies over the sensitivity region; the Imbens--Manski interval
adds finite-sample uncertainty. If the NIE set (and CI) lie entirely above 1, the conclusion
of positive mediation is **robust to differential misclassification** of the mediator --- a
statement the naive point estimate cannot support. Widen `region` to see how the bounds
respond to less certainty about the error mechanism.

> Note: `gesthtn` is a 5,000-row random sample shipped for illustration. A larger sample
> tightens the Imbens--Manski interval; see
> `system.file("scripts", "prepare_gesthtn.R", package = "medrobust")` to regenerate at any size.
