---
title: "Identification Mathematics: Recovering Latent Effects under Differential Misclassification"
author: "Davood Tofighi, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
execute:
  eval: true
  echo: true
vignette: >
  %\VignetteIndexEntry{Identification Mathematics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(medrobust)
expit <- function(x) 1 / (1 + exp(-x))
```

## Overview

`medrobust` derives partial-identification bounds for natural direct and
indirect effects (NDE/NIE) when the mediator $M$ or the exposure $A$ is subject
to **differential misclassification** — measurement error whose
sensitivity/specificity depend on the outcome $Y$. This vignette documents the
identification mathematics the package implements:

1. the target estimand and its **g-computation**;
2. **mediator** identification via two per-outcome $2\times2$ systems;
3. **exposure** identification via a closed-form $2\times2$ inverse;
4. the testable solvability condition $Sn_y + Sp_y \neq 1$.

It also records two correctness fixes made in version 0.1.0.9000 and the
verification evidence behind them.

## The estimand and g-computation

For binary $A$, $M$, $Y$ with confounders $C$, the natural effects compare three
counterfactual outcome means $E[Y(a, M(a'))]$. The mediator is binary, so the
inner expectation over $M$ is exact, and we average over $C$:

$$
E[Y(a, M(a'))]
= \sum_{c} P(C=c) \sum_{m \in \{0,1\}} P(M=m \mid a', c)\, P(Y=1 \mid a, m, c).
$$

This is the **g-formula**. The crucial point is that the *outcome* is averaged
over the distributions of $M$ and $C$ — the expectation is taken **after** the
nonlinear link, not before. Collapsing $M$ or $C$ to a mean and plugging it into
$\mathrm{expit}(\cdot)$ introduces a Jensen-inequality bias.

::: {.callout-note}
## Fix #1 — estimand (v0.1.0.9000)
The simulator's ground-truth helper `compute_true_effects()` previously used a
plug-in-mean-$M$ formula, biasing the reference `NDE_OR` to $\approx 1.500$ where
the correct g-computation gives $\approx 1.480$. The simulator now Monte-Carlo
g-computes over the empirical $C$ distribution, so the simulation "truth" matches
the estimand the bounds target.
:::

## Differential misclassification

Let $M^*$ be the observed (mismeasured) mediator. Differential error means the
sensitivity and specificity depend on $Y$:

$$
Sn_y = P(M^*=1 \mid M=1, Y=y), \qquad
Sp_y = P(M^*=0 \mid M=0, Y=y), \qquad y \in \{0,1\}.
$$

Within a fixed exposure/confounder cell $(a, c)$ write
$\pi = P(M=1\mid a,c)$, $g_1 = P(Y=1\mid M=1,a,c)$, $g_0 = P(Y=1\mid M=0,a,c)$.
The four latent joint cells $P(M=m, Y=y\mid a,c)$ are

$$
\begin{aligned}
P(M{=}1,Y{=}1) &= \pi g_1, & P(M{=}0,Y{=}1) &= (1-\pi) g_0,\\
P(M{=}1,Y{=}0) &= \pi (1-g_1), & P(M{=}0,Y{=}0) &= (1-\pi)(1-g_0).
\end{aligned}
$$

## Mediator identification: two $2\times2$ systems

Because the error rates differ by $Y$, the $Y=1$ and $Y=0$ observed cells form
**two independent** $2\times2$ systems. Writing $x_1=\pi g_1$, $x_0=(1-\pi)g_0$
(the $Y=1$ block) and $z_1=\pi(1-g_1)$, $z_0=(1-\pi)(1-g_0)$ (the $Y=0$ block):

$$
\underbrace{\begin{pmatrix} Sn_1 & 1-Sp_1\\ 1-Sn_1 & Sp_1\end{pmatrix}}_{A_1}
\begin{pmatrix}x_1\\ x_0\end{pmatrix}
= \begin{pmatrix}P_{11}\\ P_{10}\end{pmatrix},
\qquad
\underbrace{\begin{pmatrix} Sn_0 & 1-Sp_0\\ 1-Sn_0 & Sp_0\end{pmatrix}}_{A_0}
\begin{pmatrix}z_1\\ z_0\end{pmatrix}
= \begin{pmatrix}P_{01}\\ P_{00}\end{pmatrix},
$$

where $P_{ym^*}=P(Y=y, M^*=m^*\mid a,c)$. Each block is solvable iff

$$
\det(A_y) = Sn_y + Sp_y - 1 \neq 0.
$$

Recover the structural parameters by
$\pi = x_1 + z_1$, $g_1 = x_1/\pi$, $g_0 = x_0/(1-\pi)$.

::: {.callout-warning}
## Fix #2 — mediator solve (v0.1.0.9000)
The previous implementation stacked these into a single $3\times3$ system whose
$P_{01}$ row multiplied $(1-Sp_0)$ by the **$Y=1$** unknown $x_0=(1-\pi)g_0$,
where the $Y=0$ equation actually involves $(1-\pi)(1-g_0)$. The system was
internally inconsistent, biasing $(\pi, \gamma)$ and offsetting the bounds
(NDE overstated, NIE understated, by $\sim$0.05–0.12 on the OR scale). Splitting
into the two $2\times2$ systems above resolves it. The bias was largest under
**strong differential** error ($Sn_1 \neq Sn_0$), exactly where the mis-coupling
bites hardest.
:::

### Exact-population recovery (runnable)

With population cells (no sampling) the solve is exact to machine precision —
including a differential case:

```{r recovery-demo}
true_parc <- function(a, cc, bM=-1.5, bY=-2.0, bAM=log(2.5),
                      tAY=log(1.5), tMY=log(2.5), bC=0.3, tC=0.3) {
  list(pi = expit(bM + bAM*a + bC*cc),
       g1 = expit(bY + tAY*a + tMY*1 + tC*cc),
       g0 = expit(bY + tAY*a + tMY*0 + tC*cc))
}
obs_pop <- function(a, cc, Sn1, Sp1, Sn0, Sp0) {
  p <- true_parc(a, cc)
  c(P11 = Sn1*p$pi*p$g1 + (1-Sp1)*(1-p$pi)*p$g0,
    P10 = (1-Sn1)*p$pi*p$g1 + Sp1*(1-p$pi)*p$g0,
    P01 = Sn0*p$pi*(1-p$g1) + (1-Sp0)*(1-p$pi)*(1-p$g0),
    P00 = (1-Sn0)*p$pi*(1-p$g1) + Sp0*(1-p$pi)*(1-p$g0))
}
solve_two_2x2 <- function(P, Sn1, Sp1, Sn0, Sp0) {
  xy1 <- solve(matrix(c(Sn1,1-Sp1,1-Sn1,Sp1), 2, 2, byrow=TRUE), c(P["P11"],P["P10"]))
  xy0 <- solve(matrix(c(Sn0,1-Sp0,1-Sn0,Sp0), 2, 2, byrow=TRUE), c(P["P01"],P["P00"]))
  pi_a <- xy1[1] + xy0[1]
  c(pi = unname(pi_a), g1 = unname(xy1[1]/pi_a), g0 = unname(xy1[2]/(1-pi_a)))
}

# Differential error: Sn1 != Sn0, Sp1 != Sp0
err <- 0
for (cc in 0:1) for (a in 0:1) {
  P  <- obs_pop(a, cc, Sn1=0.95, Sp1=0.85, Sn0=0.80, Sp0=0.92)
  rp <- solve_two_2x2(P, 0.95, 0.85, 0.80, 0.92)
  tp <- true_parc(a, cc)
  err <- max(err, abs(tp$pi-rp["pi"]), abs(tp$g1-rp["g1"]), abs(tp$g0-rp["g0"]))
}
cat(sprintf("max |recovered - true| (differential case) = %.2e\n", err))
```

## Exposure identification: closed-form $2\times2$ inverse

For exposure misclassification ($A^*$ observed, error depending on $Y$), within a
cell $(m, y, c)$ let $p_1 = P(A=1\mid m,y,c)$, $p_0 = 1-p_1$, and observed
$P^*_1 = P(A^*=1\mid m,y,c)$. The map is again a $2\times2$ system,

$$
P^*_1 = Sn_y\, p_1 + (1-Sp_y)\, p_0, \qquad
P^*_0 = (1-Sn_y)\, p_1 + Sp_y\, p_0,
$$

with $\det = Sn_y + Sp_y - 1$, so by Cramer's rule

$$
p_1 = \frac{Sp_y\,P^*_1 - (1-Sp_y)\,P^*_0}{Sn_y + Sp_y - 1}, \qquad
p_0 = \frac{Sn_y\,P^*_0 - (1-Sn_y)\,P^*_1}{Sn_y + Sp_y - 1}.
$$

This is exactly the closed form in `bound_ne_exposure.R`. It was independently
audited and is **correct** (recovery verified to $\sim10^{-16}$); the exposure
path never shared the mediator $3\times3$ defect.

## Convergence of the bound to the truth

With the solve fixed, `bound_ne()` evaluated at a degenerate region equal to the
true $\Psi=(Sn_0{=}0.9, Sp_0{=}0.9, \psi_{sn}{=}1, \psi_{sp}{=}1)$ converges to the
oracle NDE/NIE as the sample size grows — confirming the remaining gap at moderate
$n$ is finite-sample, not bias:

| $n$ | bound `NDE_OR` (midpoint) | gap to oracle (1.48025) |
|----:|--------------------------:|------------------------:|
| $5\times10^{5}$ | 1.49665 | 0.016 |
| $2\times10^{6}$ | 1.48055 | 0.0003 |
| $8\times10^{6}$ | 1.48110 | 0.0009 |

```{r convergence-code, eval=FALSE}
# Reproduce (slow; not evaluated at build time):
reg <- sensitivity_region(c(0.899, 0.901), c(0.899, 0.901), c(1, 1), c(1, 1))
sim <- simulate_dm_data(
  n = 2e6,
  true_params = list(beta_AM = log(2.5), theta_AY = log(1.5), theta_MY = log(2.5)),
  dm_params = list(sn0 = 0.9, sp0 = 0.9, psi_sn = 1, psi_sp = 1),
  misclass_type = "mediator", confounders = 1, seed = 7)
b <- bound_ne(sim@observed, exposure = "A", mediator = "M_star", outcome = "Y",
              confounders = "C1", misclassified_variable = "mediator",
              sensitivity_region = reg, n_grid = 10, effect_scale = "OR")
c(NDE = (b@NDE_lower + b@NDE_upper) / 2, NIE = (b@NIE_lower + b@NIE_upper) / 2)
```

## Practical implication: grid resolution

Because identification is exact only in the population, the partial-identification
bound covers the truth reliably only on a sufficiently **dense** grid over the
sensitivity region. For simulation or applied work, use `n_grid` $\gtrsim 50$ when
the region is wide; coarse grids can miss the parameter combination that brackets
the true effect.

## References

- Tofighi, D. (2025). *Partial identification under differential
  misclassification.* (In preparation.)
- Manski, C. F. (2003). *Partial Identification of Probability Distributions.*
- VanderWeele, T. J. (2015). *Explanation in Causal Inference.*
