---
title: "Getting Started with medrobust"
author: "Davood Tofighi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with medrobust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `medrobust` package provides tools for conducting sensitivity analysis for causal mediation effects when the exposure or mediator is measured with **differential misclassification**. This vignette introduces the basic workflow.

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("data-wise/medrobust")
```

## Basic Example

```{r eval=FALSE}
library(medrobust)

# Load example data
data("arsenic_synthetic")

# Define sensitivity region
sens_region <- list(
  sn0_range = c(0.80, 0.90),
  sp0_range = c(0.80, 0.90),
  psi_sn_range = c(1.0, 2.0),
  psi_sp_range = c(1.0, 1.0)
)

# Compute bounds
bounds <- bound_ne(
  data = arsenic_synthetic,
  exposure = "A_star",
  mediator = "M",
  outcome = "Y",
  confounders = c("age", "smoking", "alcohol"),
  misclassified_variable = "exposure",
  sensitivity_region = sens_region
)

# View results
print(bounds)
```

## Next Steps

See the other vignettes for detailed examples:
- Mediator misclassification
- Exposure misclassification
- Interpretation guide
