---
title: "Methodology: Partial Identification Under Differential Misclassification"
author: "Davood Tofighi, Ph.D."
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
execute:
  eval: false
  echo: true
vignette: >
  %\VignetteIndexEntry{Methodology}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`medrobust` provides bounds for natural direct and indirect effects (NDE/NIE) when
the **exposure** or **mediator** is measured with *differential misclassification* —
error whose sensitivity and specificity depend on the outcome $Y$. This vignette
gives a self-contained method summary. For full derivations and convergence
diagnostics, see the companion
[Identification Mathematics](identification-math.html) vignette.

## Causal Framework

Consider binary exposure $A$, binary mediator $M$, binary outcome $Y$, and
pre-exposure confounders $C$. Under the standard potential-outcomes assumptions
(no unmeasured confounding, consistency, positivity, cross-world independence), the
natural effects compare counterfactual outcome means $E[Y(a, M(a'))]$:

$$
\text{NDE}: \quad E[Y(1,M(0))] - E[Y(0,M(0))],
$$
$$
\text{NIE}: \quad E[Y(1,M(1))] - E[Y(1,M(0))].
$$

These are identified from the observed data via the **g-formula**:

$$
E[Y(a, M(a'))]
= \sum_c P(C{=}c)
  \sum_{m} P(M{=}m \mid a', c)\, P(Y{=}1 \mid a, m, c).
$$

The key implementation point: the outcome expectation is taken *after* averaging
over $M$ and $C$, not by plugging in $E[M]$. Jensen-inequality bias would otherwise
inflate the NDE and deflate the NIE.

## Differential Misclassification Model

In practice, either the exposure or the mediator may be observed with error.
`medrobust` handles two scenarios.

### Mediator misclassification

The observed mediator $M^*$ satisfies

$$
Sn_y = P(M^*{=}1 \mid M{=}1, Y{=}y), \qquad
Sp_y = P(M^*{=}0 \mid M{=}0, Y{=}y), \qquad y \in \{0,1\}.
$$

When $Sn_0 = Sn_1$ and $Sp_0 = Sp_1$ the error is non-differential and classical
methods apply. `medrobust` targets the general case $Sn_y \neq Sn_{1-y}$ or
$Sp_y \neq Sp_{1-y}$, parameterized via odds ratios
$\psi_{sn} = (Sn_1/Sn_0)/((1-Sn_1)/(1-Sn_0))$ and $\psi_{sp}$ analogously.

### Exposure misclassification

The observed exposure $A^*$ satisfies the same structure with $A$ in place of $M$.
The four user-facing sensitivity parameters are the same: $sn_0, sp_0, \psi_{sn},
\psi_{sp}$.

## Identification Strategy

Because sensitivity and specificity differ by outcome stratum, the observed joint
distribution (in each $(A, C)$ or $(M, C)$ cell) decomposes into **two independent
$2 \times 2$ linear systems** — one for $Y=1$, one for $Y=0$. Each system is
invertible if and only if

$$
\det = Sn_y + Sp_y - 1 \neq 0.
$$

Inverting these systems recovers the latent true probabilities $P(M=m, Y=y \mid a, c)$
(mediator scenario) or $P(A=a \mid m, y, c)$ (exposure scenario), which are then
passed to the g-formula to compute NDE and NIE. For the **exposure** path the inverse
has a closed form (Cramér's rule); for the **mediator** path the package solves the
two $2 \times 2$ systems per cell.

At the true sensitivity parameters, population recovery is exact to machine precision
($<10^{-16}$); at finite $n$ the bound converges to the truth as $n \to \infty$.

## Testable Implications and Falsification

The model implies constraints on the observed data. For each misclassification
parameter set $\Psi = (sn_0, sp_0, \psi_{sn}, \psi_{sp})$, `medrobust` checks
whether the recovered latent probabilities are valid (all probabilities in $[0,1]$,
all conditional probabilities well-defined). Parameter sets that fail these checks
are **falsified** — they are empirically incompatible with the data. The falsified
region is a data-driven exclusion that typically shrinks the sensitivity space
substantially.

`check_compatibility()` exposes this falsification test directly:

```{r check-compat-example}
library(medrobust)
data("gesthtn")

compat <- check_compatibility(
  data = gesthtn,
  exposure = "A",
  mediator = "M_star",
  outcome = "Y",
  confounders = c("C1", "C2"),
  misclassified_variable = "mediator",
  sn0 = 0.90, sp0 = 0.85, psi_sn = 1.5, psi_sp = 1.2
)
print(compat)
```

## Bounds Construction

For a **sensitivity region** — a user-specified hyper-rectangle
$[sn_0^{\min}, sn_0^{\max}] \times [sp_0^{\min}, sp_0^{\max}] \times
[\psi_{sn}^{\min}, \psi_{sn}^{\max}] \times [\psi_{sp}^{\min}, \psi_{sp}^{\max}]$
— `medrobust` evaluates the effect at a grid of parameter sets (Latin Hypercube
Sampling by default) and computes:

$$
\hat L = \min_{\Psi \in \mathcal{F}} \widehat{\text{effect}}(\Psi), \qquad
\hat U = \max_{\Psi \in \mathcal{F}} \widehat{\text{effect}}(\Psi),
$$

where $\mathcal{F}$ is the non-falsified subset of the grid. The resulting interval
$[\hat L, \hat U]$ is the **partial-identification bound**.

## Confidence Intervals

The estimated bound $[\hat L, \hat U]$ is consistent but is **not a confidence set
for the true effect**: when the identified interval is narrow relative to endpoint
sampling uncertainty, naive coverage is well below the nominal level. The
Imbens–Manski (2004) construction corrects this by widening each endpoint by its
standard error:

$$
\bigl[\hat L - c\,\widehat{\mathrm{SE}}(\hat L),\;
       \hat U + c\,\widehat{\mathrm{SE}}(\hat U)\bigr],
$$

where $c$ interpolates between the one-sided and two-sided critical values.
`bound_ci()` implements this; the endpoint SEs are obtained by re-evaluating the
effect at the fixed optimal $\Psi$ on bootstrap resamples (no full grid search per
resample). Empirically this restores approximately nominal coverage across both
misclassification paths, multiple $n$, and multiple effect scales (OR, RR, RD).

```{r bound-ci-example}
bounds <- bound_ne(
  data = gesthtn,
  exposure = "A",
  mediator = "M_star",
  outcome = "Y",
  confounders = c("C1", "C2"),
  misclassified_variable = "mediator",
  sensitivity_region = sensitivity_region(
    sn0_range = c(0.80, 0.95),
    sp0_range = c(0.80, 0.95),
    psi_sn_range = c(1.0, 2.0),
    psi_sp_range = c(1.0, 2.0)
  ),
  n_grid = 50
)

ci <- bound_ci(
  bounds,
  data = gesthtn,
  exposure = "A",
  mediator = "M_star",
  outcome = "Y",
  confounders = c("C1", "C2"),
  misclassified_variable = "mediator",
  n_boot = 500L,
  level = 0.95
)
print(ci)
```

## Assumptions and Scope

`medrobust` makes the following assumptions:

- **No unmeasured confounding** between $(A, M, Y)$ after conditioning on $C$.
- **Consistency** and **positivity** in the usual potential-outcomes sense.
- **Cross-world independence**: $M(a') \perp Y(a, m) \mid C$ (required for NDE/NIE
  identification even without measurement error).
- The **misclassification model** is correctly specified: error depends only on $Y$
  (differential), not on other unmeasured variables.
- The sensitivity parameters lie in the user-supplied region. The bounds are valid
  for *every* $\Psi$ in that region simultaneously; if the true $\Psi$ lies outside
  the region, the bounds may not contain the truth.

## References

- Tofighi, D. (2025). Partial identification of causal mediation effects under
  differential misclassification. *Manuscript in preparation.*
- Imbens, G. W., & Manski, C. F. (2004). Confidence intervals for partially
  identified parameters. *Econometrica*, **72**(6), 1845–1857.
- Manski, C. F. (2003). *Partial Identification of Probability Distributions.*
  Springer.
- VanderWeele, T. J. (2015). *Explanation in Causal Inference.* Oxford University
  Press.
